Program Mutation is Just Program Transformation: \\ A Universal Approach to Mutant Generation

### 2023-02-28 paper/main.tex
describe the basic approach

>>>>

The key insight of this paper is that \emph{most mutation operators
  proposed in the literature do not in fact require parsing of source
  code.}  Consider, for example, one of the most commonly used
mutation operators, replacement of arithmetic operations.  We do not
need to parse a program containing the string {\tt ``x + y''} in order to
replace that string with {\tt ``x - y''}.  Instead, this change is
guaranteed to be effected if we simply search the program, represented
as a string, for all occurrences of  {\tt ``+''}, and produce for each
such occurrence, a mutant where tha character is replaced by {\tt
  ``-''}.  We can do the same for each pair of arithmetic operations,
and thus trivially implement a large fraction of the classic Mothra~\cite{offutt1996experimental}
mutation operators.

The obvious objection to this simplistic approach is that unless we
parse the code to identify arithmetic expressions, some of these
replacements will be invalid.  E.g., in a C++ program we will produce
mutants like {\tt x-+} replacing {\tt x++}.  However, most of these
instances can be avoided by slightly more judicious choice of search
target: instances of {\tt ``+''} where the preceeding or following
character is not another {\tt ``+''}, for instance.

Of course, in some cases this will still leave us with invalid
mutants.  However, rejecting such mutants is easy: the truly invalid
mutants will fail to compile.  Of course, we must pay the price of
checking each mutant to make sure that it does compile, at generation
time.

However, consider what has been gained by paying this cost.  We have
replaced a complex implementation, that requires parsing source code,
with an implementation that only requires string search-and-replace.
One can imagine an implementation of the above approach to arithmetic
operation replacement being implemented in as little as 10-20 lines of
Python code, including a call to a compiler to check for invalid
mutants.  Moreover, defining new mutation operators is not a matter of
understanding a representation of a parsed program, but is essentially
a matter of providing two strings, where the second is a proposed
replacement for the first.  Instead of code for arithmetic operator
replacement, we can imagine the above hypothetical Python program
reading in a file of thus expressed mutation operators, and
implementing an unbounded number of operators in a minimalist fashion.

Of course, simple string replacement is not general enough to handle
many interesting mutations.  Consider the classic case of statement
deletion, one of the most widely used and important mutation
operators.  What string can we replace with another string in order to
delete a statement in a C program?  We would need an ``operator'' for
each possible C statement.  However, if we allow the use of
\emph{regular expression match and replace pairs} in place of simple
strings, we can easily express this operator, with some precision:

\begin{verbatim}
(^\s*)(\S+[^{}]+.*)\n ==> \1/*\2*/\n
\end{verbatim}

In Python regular expression syntax, this expresses the replacement of
strings with a possibly-empty amount of white space followed by a
character pattern indicating a line of source code, with the same
source code, preserving indentation, turned into a comment.

\subsection{Hierarchies of Operator Applicability}
